\dppTitle

### 2014-06-16 manuscripts/dpp-msbayes/dpp-msbayes-figures.tex
Final edits to dpp ms in proofing stage.

%% Empirical results
\mFigure{../../empirical-analyses/plots/philippines-psi.pdf}{
    \ifbmc{\csentence{Estimated number of divergence events for 22 taxa from
    the Philippines.}}{}
    The (A--E) posterior and (F--J) prior probabilities of the number of
    divergence events (\divTimeNum) when the data of the 22 pairs of taxa from
    the Philippines are analyzed under the five models indicated at the top of
    each column of plots (Table~\ref{tabEmpiricalModels}).
    The average prior probability of an (K--O) unordered and (P--T)
    ordered model of divergence (\divTimeIndexVector) with \divTimeNum
    divergence-time parameters is also shown.
    The posterior median of the dispersion index of divergence times
    (\divTimeDispersion) is also given for each model, followed by the 95\%
    highest posterior density interval in parantheses.
}{figPhilippines}

\widthFigure{0.4}{../../empirical-analyses/plots/negros-panay-psi.pdf}{listformat=figList}{
    \ifbmc{\csentence{Estimated number of divergence events for 9 taxa from the
    Philippines.}}{}
    The posterior probabilities of the number of divergence events,
    \divTimeNum, when the data of the 9 pairs of taxa from Negros and Panay
    Islands are analyzed under the DPP model that samples over (A) unordered
    and (B) ordered models of divergence (Table~\ref{tabEmpiricalModels}).
    Both models share the same (C) prior probability of the number of
    divergence events, and the average prior probability of an (D)
    unordered and (E) ordered model of divergence (\divTimeIndexVector) with
    \divTimeNum divergence-time parameters.
    The posterior median of the dispersion index of divergence times
    (\divTimeDispersion) is also given for each model, followed by the 95\%
    highest posterior density interval in parantheses.
}{figNegrosPanay}
\clearpage

>>>>

%% Empirical results
\mFigure{../../empirical-analyses/plots/philippines-psi.pdf}{
    \ifbmc{\csentence{Estimated number of divergence events for 22 taxa from
    the Philippines.}}{}
    The (A--E) posterior and (F--J) prior probabilities of the number of
    divergence events (\divTimeNum) when the data of the 22 pairs of taxa from
    the Philippines are analyzed under the five models indicated at the top of
    each column of plots (Table~\ref{tabEmpiricalModels}).
    The average prior probability of an (K--O) unordered and (P--T)
    ordered model of divergence (\divTimeIndexVector) with \divTimeNum
    divergence-time parameters is also shown.
    The posterior median of the dispersion index of divergence times
    (\divTimeDispersion) is also given for each model, followed by the 95\%
    highest posterior density interval in parentheses.
}{figPhilippines}

\widthFigure{0.4}{../../empirical-analyses/plots/negros-panay-psi.pdf}{listformat=figList}{
    \ifbmc{\csentence{Estimated number of divergence events for 9 taxa from the
    Philippines.}}{}
    The posterior probabilities of the number of divergence events,
    \divTimeNum, when the data of the 9 pairs of taxa from Negros and Panay
    Islands are analyzed under the DPP model that samples over (A) unordered
    and (B) ordered models of divergence (Table~\ref{tabEmpiricalModels}).
    Both models share the same (C) prior probability of the number of
    divergence events, and the average prior probability of an (D)
    unordered and (E) ordered model of divergence (\divTimeIndexVector) with
    \divTimeNum divergence-time parameters.
    The posterior median of the dispersion index of divergence times
    (\divTimeDispersion) is also given for each model, followed by the 95\%
    highest posterior density interval in parentheses.
}{figNegrosPanay}
\clearpage

### 2014-06-16 manuscripts/dpp-msbayes/dpp-msbayes-main.tex
Final edits to dpp ms in proofing stage.

\subsection{Differences between this model and the original \texttt{msBayes} model}
\subsubsection{The prior on divergence models}
One of the key differences between my model and that of \msb \cite{Huang2011}
is the prior distribution on divergence models.
As discussed in Oaks et al.\ \cite{Oaks2012}, in \msb the prior used for
\divTimeIndexVector is a combination of a discrete uniform prior over the
possible number of divergence events \divTimeNum from 1 to \npairs{} with a
multinomial distribution on the number of times each index of \divTimeVector
appears in \divTimeIndexVector, with the constraint that all \divTime{}
parameters are represented at least once (see Equation 2 of \cite{Oaks2012}).
I denote this prior used by \msb as $\divTimeIndexVector \sim \priorOld$.
Oaks et al.\ \cite{Oaks2012} discuss how placing a uniform prior over the
number of divergence parameters (denoted \divTimeNum here, and as \numt{} in
\cite{Huang2011}) imposes an U-shaped prior over divergence
models (\divTimeIndexVector; see
Figure 5B of \cite{Oaks2012}).
To avoid this, I place priors directly on the sample space of divergence
models, thus eliminating the parameter \numt{} from the model.
I introduce two priors on divergence models:
(1) a prior that is uniform over all unordered divergence models, and
(2) a Dirichlet-process prior on all ordered divergence models.
The latter provides an investigator with a great deal of flexibility in
expressing their prior beliefs about models of divergence.

@@

\subsection{Power analyses: Model choice}
The modifications of the \msb model decrease the method's bias toward clustered
divergences when applied to data generated under random divergence times
(Figure~\ref{figPowerPsiOld4} \& 
S\labelcref{figPowerPsiOld,figPowerPsiUniform,figPowerPsiExp}).
The \modelOld model performs the worst of the four models across
all three series of data-generating models, inferring a single divergence event across
most of the 18,000 simulations
(Figure~\ref{figPowerPsiOld4}A--D \& plots A--F of Figures
S\labelcref{figPowerPsiOld,figPowerPsiUniform,figPowerPsiExp}).
Importantly, the \modelOld tends to strongly support these estimates
of one divergence across most of the simulations
(Figure~\ref{figPowerPsiProbOld4}A--D \& plots A--F of Figures
S\labelcref{figPowerPsiProbOld,figPowerPsiProbUniform,figPowerPsiProbExp}).
The \modelDPP model also prefers the one-divergence model 
when divergences are random over narrow windows of time, but
performs much better when divergences are random over a timescale
of 1--2 coalescent units
(Figure~\ref{figPowerPsiOld4}M--P \& plots S--X of Figures
S\labelcref{figPowerPsiOld,figPowerPsiUniform,figPowerPsiExp}).
However, even when \modelDPP infers the one-divergence model over narrow
timescales, the posterior probability support is always low
(Figure~\ref{figPowerPsiProbOld4}M--P \& plots S--X of Figures
S\labelcref{figPowerPsiProbOld,figPowerPsiProbUniform,figPowerPsiProbExp}).
% Even when the simulated data are identically distributed as the \modelOld model
% (except for the divergence times) the \modelOld is still strongly biased
% towards a single event (Figure \ref{figPowerPsiOldOld}), whereas the \modelDPP
% model is the least likely to infer a single event when applied to these data
% (Figure \ref{figPowerPsiOldDPP}).
The \modelUniform model never infers the one-divergence model in any of the
simulation replicates but still tends to infer relatively few (4--6) divergence
events when divergences are random over longer periods 
(Figure~\ref{figPowerPsiOld4} I--L \& plots M--R of Figures
S\labelcref{figPowerPsiOld,figPowerPsiUniform,figPowerPsiExp}).
Using exponential priors on divergence-time and demographic parameters does
increase the power of the \modelUshaped model compared to \modelOld across all
three series of data models, but the U-shaped prior still prevents the model
from performing as well as the \modelDPP and \modelUniform models 
(Figure~\ref{figPowerPsiOld4} \& 
S\labelcref{figPowerPsiOld,figPowerPsiUniform,figPowerPsiExp}).

@@

As mentioned above, the increased power of the new models is also evident when
looking at the estimated posterior probability of the one-divergence model
across the power analyses 
(Figure~\ref{figPowerPsiProbOld4} \& 
S\labelcref{figPowerPsiProbOld,figPowerPsiUniform,figPowerPsiExp}).
The \modelDPP and \modelUniform models estimate low posterior probability of
$\divTimeNum = 1$ across all of the simulations.
This is in contrast to the \modelOld model, which infers high posterior probabilities of a single
divergence for most replicates across all simulations
(Figure~\ref{figPowerPsiProbOld4} \& 
S\labelcref{figPowerPsiProbOld,figPowerPsiUniform,figPowerPsiExp}).
The exponential priors on divergence-time and demographic parameters (model
\modelUshaped) result in lower estimates of the probability of one divergence
when compared to \modelOld, but higher estimates when compared to \modelUniform
and \modelDPP
(Figure~\ref{figPowerPsiProbOld4} \& 
S\labelcref{figPowerPsiProbOld,figPowerPsiUniform,figPowerPsiExp}).
The \modelDPP and \modelUniform models do frequently support the one-divergence
model according to a Bayes factor criterion of greater than 10, but still
less frequently than the \modelOld model.
This result is not surprising given the extremely small prior probability of
the one-divergence model under the \modelDPP and \modelUniform models (i.e.,
very few posterior samples of the one-divergence model will result in a
large Bayes factor under these models).
However, the small posterior probability of the one-divergence model estimated
under \modelDPP and \modelUniform should prevent an investigator from
overinterpreting the Bayes factor as strong support for clustered divergences.

Lastly, when looking at the estimated posterior probability of
\divTimeDispersion being consistent with one shared divergence
($p(\divTimeDispersion < 0.01 | \ssSpace)$), I find the same pattern of model
behavior, with \modelDPP and \modelUniform inferring low probabilities across
all simulations, \modelOld inferring high probabilities, and \modelUshaped
inferring intermediate values
(Figure~\ref{figPowerOmegaProbOld4} \& 
S\labelcref{figPowerOmegaProbOld,figPowerOmegaUniform,figPowerOmegaExp}).

@@

Figure \ref{figNegrosPanay} shows the estimated posterior probability
distribution over the number of divergence events when the data from the
nine-taxon pairs from the Islands of Negros and Panay are analyzed under DPP
models that sample over unordered (\npModelDPP) and ordered
(\npModelDPPOrdered) models of divergence.
The results are similar under both models and, again, yield a large amount of
uncertainty about the number of divergence events that is similar to the prior
uncertainty.
\change{
The small difference between the results of the \npModelDPP and
\npModelDPPOrdered models is consistent across multiple analyses, and thus
could be due to error introduced to the \npModelDPP model by the invalid
shuffling of the summary statitic vectors.
}
Both models estimate a similar set of 10 unordered divergence models with the
highest posterior probability
(Figures~S\labelcref{figDivModelsNP,figDivModelsNPOrdered}).

>>>>

\subsection{Differences between this model and the original \texttt{msBayes} model}
\subsubsection{The prior on divergence models}
One of the key differences between my model and that of \msb \cite{Huang2011}
is the prior distribution on divergence models.
As discussed in Oaks et al.\ \cite{Oaks2012}, in \msb the prior used for
\divTimeIndexVector is a combination of a discrete uniform prior over the
possible number of divergence events \divTimeNum from 1 to \npairs{} with a
multinomial distribution on the number of times each index of \divTimeVector
appears in \divTimeIndexVector, with the constraint that all \divTime{}
parameters are represented at least once (see Equation 2 of \cite{Oaks2012}).
I denote this prior used by \msb as $\divTimeIndexVector \sim \priorOld$.
Oaks et al.\ \cite{Oaks2012} discuss how placing a uniform prior over the
number of divergence parameters (denoted \divTimeNum here, and as \numt{} in
\cite{Huang2011}) imposes an ``U-shaped'' prior over divergence
models (\divTimeIndexVector; see
Figure 5B of \cite{Oaks2012}).
To avoid this, I place priors directly on the sample space of divergence
models, thus eliminating the parameter \numt{} from the model.
I introduce two priors on divergence models:
(1) a prior that is uniform over all unordered divergence models, and
(2) a Dirichlet-process prior on all ordered divergence models.
The latter provides an investigator with a great deal of flexibility in
expressing their prior beliefs about models of divergence.

@@

\subsection{Power analyses: Model choice}
The modifications of the \msb model decrease the method's bias toward clustered
divergences when applied to data generated under random divergence times
(Figure~\ref{figPowerPsiOld4} \& 
S\labelcref{figPowerPsiOld,figPowerPsiUniform,figPowerPsiExp}).
The \modelOld model performs the worst of the four models across
all three series of data-generating models, inferring a single divergence event across
most of the 18,000 simulations
(Figure~\ref{figPowerPsiOld4}A--D \& plots A--F of Figures
S\labelcref{figPowerPsiOld,figPowerPsiUniform,figPowerPsiExp}).
Importantly, the \modelOld \change{model} tends to strongly support these
estimates of one divergence across most of the simulations
(Figure~\ref{figPowerPsiProbOld4}A--D \& plots A--F of Figures
S\labelcref{figPowerPsiProbOld,figPowerPsiProbUniform,figPowerPsiProbExp}).
The \modelDPP model also prefers the one-divergence model 
when divergences are random over narrow windows of time, but
performs much better when divergences are random over a timescale
of 1--2 coalescent units
(Figure~\ref{figPowerPsiOld4}M--P \& plots S--X of Figures
S\labelcref{figPowerPsiOld,figPowerPsiUniform,figPowerPsiExp}).
However, even when \modelDPP infers the one-divergence model over narrow
timescales, the posterior probability support is always low
(Figure~\ref{figPowerPsiProbOld4}M--P \& plots S--X of Figures
S\labelcref{figPowerPsiProbOld,figPowerPsiProbUniform,figPowerPsiProbExp}).
% Even when the simulated data are identically distributed as the \modelOld model
% (except for the divergence times) the \modelOld is still strongly biased
% towards a single event (Figure \ref{figPowerPsiOldOld}), whereas the \modelDPP
% model is the least likely to infer a single event when applied to these data
% (Figure \ref{figPowerPsiOldDPP}).
The \modelUniform model never infers the one-divergence model in any of the
simulation replicates but still tends to infer relatively few (4--6) divergence
events when divergences are random over longer periods 
(Figure~\ref{figPowerPsiOld4} I--L \& plots M--R of Figures
S\labelcref{figPowerPsiOld,figPowerPsiUniform,figPowerPsiExp}).
Using exponential priors on divergence-time and demographic parameters does
increase the power of the \modelUshaped model compared to \modelOld across all
three series of data models, but the U-shaped prior still prevents the model
from performing as well as the \modelDPP and \modelUniform models 
(Figure~\ref{figPowerPsiOld4} \& 
S\labelcref{figPowerPsiOld,figPowerPsiUniform,figPowerPsiExp}).

@@

As mentioned above, the increased power of the new models is also evident when
looking at the estimated posterior probability of the one-divergence model
across the power analyses 
(Figure~\ref{figPowerPsiProbOld4} \& 
S\labelcref{figPowerPsiProbOld,figPowerPsiProbUniform,figPowerPsiProbExp}).
The \modelDPP and \modelUniform models estimate low posterior probability of
$\divTimeNum = 1$ across all of the simulations.
This is in contrast to the \modelOld model, which infers high posterior probabilities of a single
divergence for most replicates across all simulations
(Figure~\ref{figPowerPsiProbOld4} \& 
S\labelcref{figPowerPsiProbOld,figPowerPsiProbUniform,figPowerPsiProbExp}).
The exponential priors on divergence-time and demographic parameters (model
\modelUshaped) result in lower estimates of the probability of one divergence
when compared to \modelOld, but higher estimates when compared to \modelUniform
and \modelDPP
(Figure~\ref{figPowerPsiProbOld4} \& 
S\labelcref{figPowerPsiProbOld,figPowerPsiProbUniform,figPowerPsiProbExp}).
The \modelDPP and \modelUniform models do frequently support the one-divergence
model according to a Bayes factor criterion of greater than 10, but still
less frequently than the \modelOld model.
This result is not surprising given the extremely small prior probability of
the one-divergence model under the \modelDPP and \modelUniform models (i.e.,
very few posterior samples of the one-divergence model will result in a
large Bayes factor under these models).
However, the small posterior probability of the one-divergence model estimated
under \modelDPP and \modelUniform should prevent an investigator from
overinterpreting the Bayes factor as strong support for clustered divergences.

Lastly, when looking at the estimated posterior probability of
\divTimeDispersion being consistent with one shared divergence
($p(\divTimeDispersion < 0.01 | \ssSpace)$), I find the same pattern of model
behavior, with \modelDPP and \modelUniform inferring low probabilities across
all simulations, \modelOld inferring high probabilities, and \modelUshaped
inferring intermediate values
(Figure~\ref{figPowerOmegaProbOld4} \& 
S\labelcref{figPowerOmegaProbOld,figPowerOmegaProbUniform,figPowerOmegaProbExp}).

@@

Figure \ref{figNegrosPanay} shows the estimated posterior probability
distribution over the number of divergence events when the data from the
nine-taxon pairs from the Islands of Negros and Panay are analyzed under DPP
models that sample over unordered (\npModelDPP) and ordered
(\npModelDPPOrdered) models of divergence.
The results are similar under both models and, again, yield a large amount of
uncertainty about the number of divergence events that is similar to the prior
uncertainty.
\change{
The small difference between the results of the \npModelDPP and
\npModelDPPOrdered models is consistent across multiple analyses, and thus
could be due to error introduced to the \npModelDPP model by the invalid
shuffling of the summary statistic vectors.
}
Both models estimate a similar set of 10 unordered divergence models with the
highest posterior probability
(Figures~S\labelcref{figDivModelsNP,figDivModelsNPOrdered}).
